---
title: "QR Algorithm for Eigenvalue Computation"
author: "KHALID BOUKHIT"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: show
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 6)
library(ggplot2)
library(microbenchmark)


set.seed(42)
```

# Introduction

The **QR Algorithm** is the workhorse of numerical linear algebra for computing **all** eigenvalues of a matrix. It is the algorithm behind R's `eigen()`, MATLAB's `eig()`, and LAPACK's `xGEEV`/`xSYEV` routines. Unlike power/inverse iteration (which find one eigenvalue at a time), the QR algorithm simultaneously converges the matrix to a form from which all eigenvalues can be read off.

# Mathematical Background

## QR Decomposition

Any matrix $A \in \mathbb{R}^{n \times n}$ can be factored as $A = QR$ where $Q$ is orthogonal ($Q^\top Q = I$) and $R$ is upper triangular.

## Basic QR Iteration

Starting with $A_0 = A$:

$$A_k = Q_k R_k \quad \text{(QR decomposition)}$$
$$A_{k+1} = R_k Q_k \quad \text{(reverse multiply)}$$


$A_{k+1} = R_k Q_k = Q_k^\top (Q_k R_k) Q_k = Q_k^\top A_k Q_k$, so all $A_k$ are **similar** (same eigenvalues). The sequence $A_k$ converges to an upper-triangular (or quasi-triangular) matrix whose diagonal entries are the eigenvalues.

## Why It Converges

The QR algorithm is equivalent to **simultaneous iteration** (applying power iteration to all $n$ basis vectors at once with re-orthogonalisation). The convergence of $A_k$ to upper-triangular form is governed by the ratios $|\lambda_{i+1}/\lambda_i|$. Sub-diagonal elements $a_{i+1,i}^{(k)}$ converge to zero at rate $|\lambda_{i+1}/\lambda_i|^k$.

## Shifted QR Algorithm

Convergence is dramatically accelerated by introducing **shifts**. At each step:
$$A_k - \sigma_k I = Q_k R_k$$
$$A_{k+1} = R_k Q_k + \sigma_k I$$

The shift $\sigma_k$ is chosen to approximate an eigenvalue, driving relevant sub-diagonal elements to zero rapidly.

### Wilkinson Shift

The **Wilkinson shift** is the eigenvalue of the trailing $2 \times 2$ sub-matrix:

$$B = \begin{pmatrix} a_{n-1,n-1} & a_{n-1,n} \\ a_{n,n-1} & a_{n,n} \end{pmatrix}$$

that is closest to $a_{n,n}$. The eigenvalues of $B$ are:

$$\mu = \frac{\text{tr}(B)}{2} \pm \sqrt{\left(\frac{\text{tr}(B)}{2}\right)^2 - \det(B)}$$

The Wilkinson shift gives **globally convergent** cubic convergence for symmetric matrices.

## Deflation

When $|a_{m,m-1}^{(k)}| < \epsilon$, we accept $a_{m,m}^{(k)}$ as a converged eigenvalue and **deflate**: reduce the working matrix to the leading $(m-1) \times (m-1)$ block.

## Complexity

For an $n \times n$ matrix:

- Without preliminary reduction: $O(n^3)$ per QR step → total $O(n^4)$ 
- With Hessenberg reduction (standard practice): $O(n^2)$ per QR step → total $O(n^3)$
- With shifts: convergence in $O(n)$ total iterations → total $O(n^3)$



# Algorithm Pseudocode

## Basic QR

```
QR_BASIC(A, tol, max_iter):
    A_k ← A
    for k = 1, 2, …, max_iter:
        Q, R ← QR(A_k)
        A_k ← R · Q
        if sum(|lower-triangular(A_k)|) < tol:
            RETURN diag(A_k)   # eigenvalues
    RETURN diag(A_k)
```

## Shifted QR with Deflation

```
QR_SHIFTED(A, tol, max_iter):
    A_k ← A;  m ← n
    while m > 1:
        σ ← wilkinson_shift(A_k[m-1:m, m-1:m])
        Q, R ← QR(A_k[1:m, 1:m] − σ I)
        A_k[1:m, 1:m] ← R · Q + σ I
        if |A_k[m, m-1]| < tol:
            eigenvalue[m] ← A_k[m, m]
            m ← m − 1         # deflate
    eigenvalue[1] ← A_k[1,1]
    RETURN eigenvalues
```

# R Implementation

```{r source-impl}
###############################################################################
# QR Algorithm — Pure R Implementation
# Computes ALL eigenvalues of a square matrix via iterative QR decomposition.
###############################################################################

#' QR Decomposition (Gram-Schmidt)
#'
#' Pure R QR decomposition. Returns Q (orthogonal) and R (upper triangular).
#' @param A Numeric matrix.
#' @return List with Q and R.
qrdecomp <- function(A) {
  n <- nrow(A)
  m <- ncol(A)
  Q <- matrix(0, n, m)
  R <- matrix(0, m, m)

  for (j in seq_len(m)) {
    v <- A[, j]
    if (j > 1) {
      for (i in 1:(j - 1)) {
        R[i, j] <- sum(Q[, i] * A[, j])
        v <- v - R[i, j] * Q[, i]
      }
    }
    R[j, j] <- sqrt(sum(v^2))
    if (R[j, j] > 1e-14) {
      Q[, j] <- v / R[j, j]
    } else {
      Q[, j] <- v
    }
  }
  list(Q = Q, R = R)
}

#' Basic QR Algorithm (no shifts)
#'
#' @param A       Square numeric matrix (should be real).
#' @param tol     Convergence tolerance (sub-diagonal element magnitude).
#' @param max_iter Maximum iterations.
#' @return List with: values (eigenvalues), iterations, converged, history.
qralgorithmbasic <- function(A, tol = 1e-10, max_iter = 500) {
  n <- nrow(A)
  Ak <- A
  history <- list()

  for (k in seq_len(max_iter)) {
    decomp <- qrdecomp(Ak)
    Ak <- decomp$R %*% decomp$Q
    off_diag <- sum(abs(Ak[lower.tri(Ak)]))
    history[[k]] <- list(eigenvalues = diag(Ak), off_diag = off_diag)

    if (off_diag < tol) {
      return(list(
        values     = diag(Ak),
        matrix     = Ak,
        iterations = k,
        converged  = TRUE,
        history    = history
      ))
    }
  }
  list(
    values     = diag(Ak),
    matrix     = Ak,
    iterations = max_iter,
    converged  = FALSE,
    history    = history
  )
}

#' QR Algorithm with Wilkinson Shift
#'
#' Uses the Wilkinson shift (eigenvalue of trailing 2×2 sub-matrix closest
#' to a_{nn}) to accelerate convergence.
#' @param A       Square symmetric numeric matrix.
#' @param tol     Convergence tolerance.
#' @param max_iter Maximum iterations.
#' @return List with: values, iterations, converged, history.
qralgorithmshifted <- function(A, tol = 1e-10, max_iter = 1000) {
  n <- nrow(A)
  Ak <- A
  eigenvalues <- numeric(n)
  m <- n
  total_iter <- 0
  history <- list()

  while (m > 1) {
    for (k in seq_len(max_iter)) {
      total_iter <- total_iter + 1
      # Wilkinson shift from trailing 2×2 block
      a <- Ak[m - 1, m - 1]
      b <- Ak[m - 1, m]
      c <- Ak[m, m - 1]
      d <- Ak[m, m]
      tr <- a + d
      det_val <- a * d - b * c
      disc <- sqrt(max(tr^2 / 4 - det_val, 0))
      mu1 <- tr / 2 + disc
      mu2 <- tr / 2 - disc
      sigma <- ifelse(abs(mu1 - d) < abs(mu2 - d), mu1, mu2)

      # Shifted QR step
      decomp <- qrdecomp(Ak[1:m, 1:m] - sigma * diag(m))
      Ak[1:m, 1:m] <- decomp$R %*% decomp$Q + sigma * diag(m)

      history[[total_iter]] <- list(
        eigenvalues = diag(Ak[1:m, 1:m]),
        sub_diag    = abs(Ak[m, m - 1])
      )

      if (abs(Ak[m, m - 1]) < tol) {
        eigenvalues[m] <- Ak[m, m]
        m <- m - 1
        break
      }
    }
  }
  if (m == 1) eigenvalues[1] <- Ak[1, 1]

  list(
    values     = sort(eigenvalues, decreasing = TRUE),
    iterations = total_iter,
    converged  = TRUE,
    history    = history
  )
}
```


## Custom QR Decomposition

```{r show-qr-decomp, eval=FALSE}
qrdecomp <- function(A) {
  n <- nrow(A); m <- ncol(A)
  Q <- matrix(0, n, m); R <- matrix(0, m, m)
  for (j in seq_len(m)) {
    v <- A[, j]
    if (j > 1) {
      for (i in 1:(j-1)) {
        R[i, j] <- sum(Q[, i] * A[, j])
        v <- v - R[i, j] * Q[, i]
      }
    }
    R[j, j] <- sqrt(sum(v^2))
    Q[, j] <- if (R[j, j] > 1e-14) v / R[j, j] else v
  }
  list(Q = Q, R = R)
}
```

# Testing with Sample Matrices

## Example 1: 3×3 Symmetric — Basic QR

```{r test-basic-3x3}
A1 <- matrix(c(4, 1, 2,
               1, 3, 0,
               2, 0, 5), nrow = 3, byrow = TRUE)
cat("Matrix A1:\n"); print(A1)

res1basic <- qralgorithmbasic(A1)
cat("\nEigenvalues (basic QR):", sort(res1basic$values, decreasing = TRUE))
cat("\nIterations:", res1basic$iterations)
cat("\nBuilt-in eigen():", sort(eigen(A1)$values, decreasing = TRUE))
```

## Example 2: 3×3 Symmetric — Shifted QR

```{r test-shifted-3x3}
res1shifted <- qralgorithmshifted(A1)
cat("Eigenvalues (shifted QR):", res1shifted$values)
cat("\nIterations:", res1shifted$iterations)
```

## Example 3: 5×5 Random Symmetric

```{r test-5x5}
set.seed(42)
A3 <- matrix(rnorm(25), 5, 5)
A3 <- A3 + t(A3)

res3basic   <- qralgorithmbasic(A3)
res3shifted <- qralgorithmshifted(A3)

cat("Basic  QR eigenvalues:", round(sort(res3basic$values, decreasing = TRUE), 6))
cat("\nShifted QR eigenvalues:", round(res3shifted$values, 6))
cat("\nBuilt-in eigen():", round(sort(eigen(A3)$values, decreasing = TRUE), 6))
cat("\n\nBasic iterations:", res3basic$iterations)
cat("\nShifted iterations:", res3shifted$iterations)
```

# Comparison with Built-in `eigen()`

```{r comparison}
evalstrue <- sort(eigen(A3)$values, decreasing = TRUE)
evalsbasic <- sort(res3basic$values, decreasing = TRUE)
evalsshifted <- sort(res3shifted$values, decreasing = TRUE)

comparison <- data.frame(
  Index = 1:5,
  BuiltIn = round(evalstrue, 10),
  BasicQR = round(evalsbasic, 10),
  ShiftedQR = round(evalsshifted, 10),
  Error_Basic = abs(evalsbasic - evalstrue),
  Error_Shifted = abs(evalsshifted - evalstrue)
)
knitr::kable(comparison, digits = 12, caption = "QR Algorithm vs eigen()")
```

# Convergence Plot — Sub-Diagonal Decay

```{r convergence-sub-diag}
# Track off-diagonal norm for basic QR on A1
resconv <- qralgorithmbasic(A1, tol = 1e-15, max_iter = 200)
offdiags <- sapply(resconv$history, function(h) h$off_diag)

dfoff <- data.frame(Iteration = seq_along(offdiags), OffDiag = offdiags)
dfoff <- dfoff[dfoff$OffDiag > 0, ]

ggplot(dfoff, aes(x = Iteration, y = OffDiag)) +
  geom_line(colour = "#2c3e50", linewidth = 0.8) +
  scale_y_log10() +
  labs(title = "QR Algorithm: Sub-diagonal Norm Decay",
       x = "Iteration", y = "|Lower-triangular elements| (log scale)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

# Eigenvalue Trajectory Plot

```{r eigenvalue-trajectory}
# Track diagonal elements (eigenvalue estimates) across iterations
ndiag <- ncol(A1)
iters <- length(resconv$history)
eigtraj <- matrix(NA, iters, ndiag)
for (i in seq_len(iters)) {
  eigtraj[i, ] <- sort(resconv$history[[i]]$eigenvalues, decreasing = TRUE)
}

dftraj <- data.frame(
  Iteration = rep(1:iters, ndiag),
  Value = as.numeric(eigtraj),
  Eigenvalue = factor(rep(paste0("λ", 1:ndiag), each = iters))
)

trueevals <- sort(eigen(A1)$values, decreasing = TRUE)
dftrue <- data.frame(
  Eigenvalue = paste0("λ", 1:ndiag),
  TrueValue = trueevals
)

ggplot(dftraj, aes(x = Iteration, y = Value, colour = Eigenvalue)) +
  geom_line(linewidth = 0.7) +
  geom_hline(data = dftrue, aes(yintercept = TrueValue),
             linetype = "dashed", alpha = 0.5) +
  labs(title = "Eigenvalue Estimates During QR Iteration",
       x = "Iteration", y = "Diagonal Element Value") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

# Basic vs Shifted Iteration Count

```{r basic-vs-shifted}
sizes <- c(4, 6, 8, 10, 15, 20)
itercomparison <- data.frame(Size = integer(), Method = character(),
                              Iterations = integer())
for (n in sizes) {
  set.seed(42)
  M <- matrix(rnorm(n * n), n, n)
  M <- M + t(M)

  r_basic <- qralgorithmbasic(M, tol = 1e-10, max_iter = 2000)
  r_shift <- qralgorithmshifted(M, tol = 1e-10, max_iter = 2000)

  itercomparison <- rbind(itercomparison,
    data.frame(Size = c(n, n),
               Method = c("Basic QR", "Shifted QR"),
               Iterations = c(r_basic$iterations, r_shift$iterations)))
}

ggplot(itercomparison, aes(x = Size, y = Iterations, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Iterations to Convergence: Basic vs Shifted QR",
       x = "Matrix Size", y = "Iterations") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

# Runtime Benchmarks

```{r benchmarks}
sizes <- c(10, 25, 50, 75, 100)
benchresults <- data.frame(Size = integer(), Method = character(),
                            MedianTime_ms = numeric())

for (n in sizes) {
  set.seed(42)
  M <- matrix(rnorm(n * n), n, n)
  M <- M + t(M)

  bm <- microbenchmark(
    BasicQR      = qralgorithmbasic(M, tol = 1e-8),
    ShiftedQR    = qralgorithmshifted(M, tol = 1e-8),
    BuiltinEigen = eigen(M),
    times = 5
  )
  sm <- summary(bm)
  benchresults <- rbind(benchresults, data.frame(
    Size = rep(n, 3),
    Method = c("Basic QR", "Shifted QR", "Built-in eigen()"),
    MedianTime_ms = sm$median / 1e6
  ))
}

knitr::kable(benchresults, digits = 3, caption = "Median Runtime (ms)")
```

```{r benchmark-plot}
ggplot(benchresults, aes(x = Size, y = MedianTime_ms, colour = Method)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_y_log10() +
  labs(title = "Runtime: Basic QR vs Shifted QR vs eigen()",
       x = "Matrix Size (n)", y = "Median Time (ms, log scale)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

# Summary

| Property | Basic QR | Shifted QR |
|---|---|---|
| **Target** | All eigenvalues | All eigenvalues |
| **Convergence** | Linear ($|\lambda_{i+1}/\lambda_i|^k$) | Cubic (Wilkinson shift, symmetric) |
| **Per-step cost** | $O(n^3)$ | $O(n^3)$ (without Hessenberg reduction) |
| **Total cost** | $O(n^4)$ typical | $O(n^3)$ typical |
| **Finds eigenvectors** | Via accumulated $Q$ products | Via accumulated $Q$ products |
| **Used in practice** | No | Yes (basis of LAPACK routines) |
